\documentclass[twoside]{article}

\usepackage{ustj}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{pdflscape}

\addbibresource{mss.bib}

\newcommand{\authorname}{N.\ E.\ Davis}
\newcommand{\authorpatp}{\patp{lagrev-nocfep}}
\newcommand{\affiliation}{ZeroAttest, Groundwire Foundation}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{A Deterministic Numerical Stack}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{A Deterministic Numerical Stack}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

\begin{document}

\maketitle
\thispagestyle{firststyle}

\begin{abstract}
\end{abstract}

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{1}

\tableofcontents

\section{Introduction}

Urbit is a deterministic functional operating system whose state is a pure function of its event log \citep{Whitepaper}.  An event log replay must recover the state exactly, on any conforming runtime, indefinitely.  This is not a performance nicety; it is the property that makes the system auditable, upgradeable, and peer-verifiable.  Determinism is a load-bearing property all the way down.

Numerical computing is one of the most slippery places for determinism to leak through its abstractions.  Standard \textsc{ieee}~754 floating-point arithmetic is not associative, as an example, but the situation is much more dire.  The \textsc{ieee}~754 standard~\citep{ieee754-2019} specifies the result of each individual operation under a given rounding mode, but a real numerical program is a composition of operations, and the standard says little about composition.  A fused multiply-add computes $a\cdot b + c$ with a single rounding where two separate operations would round twice; a vectorized dot product sums partial results in an order chosen by the hardware; and a threaded \textsc{blas} reassociates a reduction differently depending on how many cores happened to be available.\footnote{Urbit computing is single-threaded which removes one possible source of nondeterminism.}  Each of these is standards-conformant while none is deterministic across hosts.

The treatment of floating-point representation in Hoon by \citet{Davis2024} established the foundation for Urbit's deterministic numerical computing stack.  This article expounds both the state of the art of Urbit's basic numerical packaging and the next layer:  dense linear algebra, where the nondeterminism of conventional \textsc{blas} and \textsc{ieee} 754 floating-point mathematics are acute and consequential.  We introduce SoftBLAS,\footnote{\href{https://github.com/urbit/SoftBLAS}{\texttt{urbit/SoftBLAS}}} a software-defined \textsc{blas} package; SoftUnum,\footnote{\href{https://github.com/sigilante/SoftUnum}{\texttt{sigilante/SoftUnum}}} a software-defined universal number (unum) package; and other components of a C-based numerical stack that is deterministic and compliant with the Hoon specification.  We further describe the ``ray'' data structure which facilitates the type dispatch for different numeric representations while preserving identity between Nock-specified atoms and C-recognized arrays.

The central conceit is architectural.  In conventional high-performance computing, C (or Fortran) takes primacy:  the C implementation is what runs, and correctness is in some sense whatever the C does.\footnote{Verification and validation apply, of course.}  Urbit inverts this relationship.  The Hoon/Nock \textsc{isa} implementation is the specification; the C jet is an optimization that is \emph{obligated} to produce the specification's exact output for every input, and should be discarded silently whenever it cannot.  Determinism then is not merely a testable, desirable property, but a mandate the Nock-side architecture cannot express the violation of.

\section{Background and Literature}

Although the solution of partial differential equations is one of the oldest significant applications of computing,\footnote{\citet{Charney1950}.} no standard library emerged for decades---unsurprising given the churn of hardware architectures, the diversity of numerical methods, the lack of a common interface definition, the rapid evolution of programming languages, and the active development of linear algebraic algorithms.\footnote{Even the size of a byte took decades to standardize.}  Given that situation, rigorous bitwise determinism was a concern of vanishingly small importance, whether cross-platform or even for serial runs on the same machine.  However, over time the advent of high-performance computing, the emergence of \textsc{blas} as a pseudo-standard, and the increasing importance of reproducibility in numerical computing have made determinism both more pressing and more achievable.

\subsection{\textsc{blas} as a pseudo-standard}

The Basic Linear Algebra Subprograms originate with Lawson, Hanson, Kincaid, and Krogh's 1979 specification of vector operations on strided arrays~\citep{lawson1979blas}, later extended to matrix-vector (Level~2) and matrix-matrix (Level~3) routines.  The levels track both chronology and
(na\"{i}ve) asymptotic cost:  Level~1 is $O(n)$, Level~2 is $O(n^2)$, Level~3 is $O(n^3)$.  \textsc{blas} is less a standard than a pseudo-standard---a widely honored set of names and calling conventions (\texttt{?axpy}, \texttt{?gemv}, \texttt{?gemm}, with type prefixes \texttt{s}, \texttt{d}, \texttt{c}, \texttt{z}) with many independent implementations.  (The original \textsc{blas} predated floating-point standardization as \textsc{ieee} 754 by several years.)  That plurality is exactly the problem for a consensus system:  two conforming \textsc{blas} libraries routinely disagree in the low bits, and the same library disagrees with itself across machines and thread counts.  Despite this fuzziness, \textsc{blas} quickly took its place as the bedrock layer of numerical and scientific computation.

\subsection{Sources of floating-point nondeterminism}

Cross-host divergence in a \textsc{blas} computation has a small number of recurring causes:

\begin{itemize}
  \item  \emph{Reduction order}.  Floating-point addition is not associative, so a dot product or a \texttt{gemm} inner loop summed in a different order yields a different rounded result; tiled and multithreaded implementations choose that order at run time.
  \item  \emph{Fused multiply-add}.  Hardware that contracts $a\cdot b + c$ to one rounding produces different bits than two roundings, and whether contraction happens is a compiler- and target-dependent decision.
  \item  \emph{Vectorization and instruction selection}.  \textsc{simd} horizontal sums, \texttt{x87} 80-bit intermediates, and differing \texttt{libm}-style kernels for fast paths all perturb the last bits.
  \item  \emph{Transcendental functions}.  The elementary functions are not specified to the bit by \textsc{ieee}~754 at all, and every platform \texttt{libm} differs in its implementation.
\end{itemize}

\citet{Davis2024} further enumerated other reasons that contemporary floating-point mathematics cannot be considered deterministic even when single-threaded on a particular hardware target and instruction set architecture.

\subsection{Reproducibility}

Numerical determinism, or for our purposes synonymously reproducibility, requires that a computation yield the same result to the bit regardless of the platform on which it is carried out.  This would not necessarily require the same algorithmic implementation in all cases (since all correct algorithms should converge on the true value given enough runway), but in practice floating-point quirks force this consideration in almost all cases.

Projects like ReproBLAS have yielded reproducible summation (independent of summation order) through
error-free transformations and pre-rounding~\citep{demmel2013reproducible}.  Intel's Conditional Numerical Reproducibility offers run-to-run consistency within a fixed code path under documented constraints~\citep{intel_cnr}.  The \textsc{plasma} and \textsc{magma} pro\-jects \citep{agullo2009plasma} took the opposite design stance for performance: dynamic, dependency-driven scheduling that deliberately does not fix reduction order, which is excellent for throughput and fatal for bit-reproducibility.  On the scalar side, correctly rounded elementary-function libraries---\textsc{cr-libm} \linebreak\citep{daramy2003crlibm}, \textsc{rlibm}~\citep{lim2021rlibm}, and the \textsc{core-math} project~\citep{coremath}---show that the transcendentals can be pinned down to the last bit, at a cost.

Urbit is distinguished by its requirement that determinism be not one option among many but a first-class feature of the OS and runtime.  This requirement is structurally enforced by the Nock \textsc{isa}--jetted runtime relationship.  Jets utilize relatively faster code paths and reductions but produce the identical result as the slow path:  the specification is authoritative.  Thus linear algebra techniques, even those following the \textsc{blas} pseudo-standard, must follow a reproducible fast path on all supported architectures.

\section{Deterministic Architecture}

A Nock computation is a pure function from noun to noun.  A \emph{jet} is a native routine registered against a named Hoon core; when the interpreter evaluates that core, it may run the jet instead of the Nock.  The iron discipline that makes this sound is simple:  the jet must produce the same noun as the Hoon, for every input, or it is a bug in the runtime.  A divergence between jet and specification is termed a \emph{jet mismatch} and is treated as a correctness failure, not a numerical tolerance.

The Hoon (or its compiled result as Nock) is therefore the specification, in the strong sense that it defines what the right answer \emph{is}.  The C jet is an accelerant with no semantic authority.  In some cases, the jet may decline:  if a native routine returns the sentinel \texttt{u3\_none}, the interpreter falls through and evaluates the Hoon directly.  The consequence for numerics is that an arithmetic kind with no jet, or with a jet that punts on some inputs, is not unsupported---it is merely slow.  The specification still runs, and the answer is still defined and still deterministic, because Hoon over exact nouns is deterministic by construction.

This crystallizes, or perhaps even inverts, the usual relationship between correctness and performance.  In conventional \textsc{hpc}, one has an optimized C or Fortran program, the optimized C arguably \emph{is} the program and its source reference implementation is documentation (to say nothing of its documentation).  There may of course be bugs in the compiled code, but the optimized program is the concrete artifact.  By contrast in Urbit, the optimized C program must justify itself against an independent authoritative definition.  Determinism thus follows from two facts:  the specification is a pure function of exact data, and the jet is contractually bit-identical to the specification.  A host without the jet computes the same bits more
slowly; a host with a divergent jet is non-conforming and must be repaired.

For our linked implementations like SoftBLAS, this means the library is never permitted to be ``approximately'' the specification. A \texttt{gemm} jet that summed its inner products in a hardware-chosen order would be a mismatch even if every individual rounding were \textsc{ieee} 754-correct, because the Hoon specification sums them in
one fixed order and the jet must reproduce \emph{that} number.\footnote{Summation will necessarily exhibit order-dependent drift at the bit level.}  Our \textsc{blas} must satisfy the constraint of determinism to be admissible as a jet at all; thus, SoftBLAS was developed on top of SoftFloat \citep{hauser_softfloat}.  This yields correctness and consistency rather than hardware-optimized performance.

\subsection{The numerical stack}
\label{sec:stack}

Urbit natively supports \textsc{ieee} 754 floating-point mathematics in its \texttt{\%base} distribution, and through the supplied numerical stack additionally supports unums (posits and quires); complex \textsc{ieee} 754; two's complement integers; and fixed-point mathematics.

The numerical libraries are layered as follows; each layer
is Hoon-normative with optional, independently verifiable C jets.

\begin{enumerate}
  % \item  \emph{Reference libraries}.  We desire to minimize the necessary libraries linked to Urbit, but the SoftBLAS deterministic linear algebra library and a SoftUnum universal number library have been developed by Urbit core developers to support a compliant external interface.
  \item  \emph{Libraries}.  Imported from \texttt{/lib} at compile time.
  \begin{enumerate}
    \item  \texttt{/lib/math}.  The basic \textsc{ieee} 754 arithmetic and transcendental functions library supports four precisions (16-bit \texttt{@rh}, 32-bit \texttt{@rs}, 64-bit \texttt{@rd}, and 128-bit \texttt{@rq}) across a battery of functions like exponentiation and logarithm, trigonometric and inverse-trigonometric functions, and rounding and comparison utilities.  Transcendental functions are implemented as Chebyshev minimax polynomials with Cody--Waite range reduction~\citep{codywaite1980}.  Jets run on SoftFloat \citep{hauser_softfloat}.  Originally composed by \patp{lagrev-nocfep} and \patp{mopfel-winrux}.
    \item \texttt{/lib/twoc}.  A two's-complement integer arithmetic package supporting scalar operations.  Jetted with an internal library against \textsc{gmp}.  Originally composed by \patp{litlep-nibbyt}.  
    \item \texttt{/lib/fixed}.  Fixed-point (Q-format) arithmetic at power-of-two total widths.  Unjetted beyond standard integer arithmetic jets because de facto integer mathematics hold.  Originally composed by \patp{lagrev-nocfep}.
    \item \texttt{/lib/unum}.  2022-standard posit arithmetic~\citep{positstd2022} at five widths (8-bit \texttt{@rpb}; 16-bit \texttt{@rph}; 32-bit \texttt{@rps}; 64-bit \texttt{@rpd}; 128-bit \texttt{@rpq}, i.e. posit8 through posit128), $es=2$ throughout, with quire accumulation.  See Section \ref{sec:verify} on oracle coverage.  Jetted with SoftUnum.  Originally composed by \patp{lagrev-nocfep}.
    \item \texttt{/lib/complex}.  Complex arithmetic over the four \textsc{ieee} 754 precisions, riding on the scalar float layer.  Originally composed by \patp{lagrev-nocfep}.
  \end{enumerate}
  \item \emph{Lagoon (Linear AlGebra in hOON)}.  A rank-$N$ array library (a \texttt{\$ray}) carrying a \texttt{\$kind} field that dispatches to the appropriate arithmetic:  \texttt{\%i754}, \texttt{\%uint}, \texttt{\%int2}, \texttt{\%unum}, \texttt{\%cplx}, \texttt{\%fixp}.  Storage is row-major and identical in memory layout to underlying runtime arrays; the \texttt{\%i754} Level-3 path is jetted through SoftBLAS and the \texttt{\%unum} path is jetted through SoftUnum.
  \item  \emph{Saloon (Scientific ALgorithms in hOON)}.  Scientific algorithms over Lagoon rays.  Types correspond to the Lagoon field.  Element-wise transcendentals are dispatched through \texttt{/lib/math}, and dense routines such as the Jacobi symmetric eigendecomposition.\footnote{Broadly speaking, Lagoon:NumPy::Saloon:SciPy.}
  \item  \emph{Maroon (MAchine leaRning in hOON)}.  An inchoate machine learning library built atop Lagoon and Saloon.
\end{enumerate}

The single most consequential design choice is Lagoon's \texttt{\$kind} dispatch with fall-through (Listing~\ref{lst:ray}).  One basic array type serves six arithmetics; adding a seventh requires only a Hoon implementation, and a jet for it is optional and separately auditable.  This is architecturally distinct from NumPy (which supplies one \texttt{dtype} per array with no in-band kind tag); from Julia (which has compile-time specialization per concrete type); and from the \textsc{apl}/J/K tradition (which supply a uniform numeric tower with no explicit kind dispatch).  Because an un-jetted kind still runs its Hoon, the specification is authoritative for every kind, jetted or not.  Two other features of \texttt{\$ray}'s design call attention:  the \textsc{msb} pinned \texttt{1} bit, which preserves otherwise-stripped leading zeroes in atoms and thus makes the atom representation exactly match the expected C memory layout of an array; and the \texttt{tail=*} arbitrary noun specialization field, which preserves jet axis addresses in case of future extension.\footnote{For \texttt{\%cplx}, \texttt{bloq} is the $\log_{2}$ of the packed-pair element width (both components combined).  Thus \texttt{@cs} (two \texttt{@rs} components) has \texttt{bloq=6}, while each component \texttt{@rs} has \texttt{bloq=5}.}

\begin{lstlisting}[style=listingblock,
                   label=lst:ray,
                   caption={Lagoon base types.  From \texttt{/sur/lagoon}.}]
+$  ray               ::  $ray:  n-dimensional array
  $:  =meta           ::  descriptor
      data=@ux        ::  data, row-major order
  ==
::
+$  prec  [a=@ b=@]   ::  fixed-point precision
+$  meta              ::  $meta:  metadata for a $ray
  $:  shape=(list @)  ::  list of dimension lengths
      =bloq           ::  logarithm of bitwidth
      =kind           ::  name of data type
      tail=*          ::  per-kind specialization data
  ==
::
+$  kind              ::  $kind:  type of array scalars
  $?  %i754           ::  IEEE 754 float
      %uint           ::  unsigned integer
      %int2           ::  2s-complement integer (/lib/twoc)
      %unum           ::  unum/posit (/lib/unum)
      %cplx           ::  complex (/lib/complex)
      %fixp           ::  fixed-point Q a.b (/lib/fixed)
  ==
::
+$  baum              ::  $baum:  ndray with metadata
  $:  =meta           ::
      data=ndray      ::
  ==
::
+$  ndray             ::  $ndray:  n-dim array as list
    $@  @             ::  single item
    (list ndray)      ::  nonempty list of children
\end{lstlisting}

\subsection{SoftBLAS and SoftUnum}

At the current time, two external numerical libraries facilitate 
\footnote{Urbit mathematics are supported internally by the \href{https://gmplib.org/}{\textsc{gnu} Multiple Precision Arithmetic Library} and \href{https://www.jhauser.us/arithmetic/SoftFloat.html}{SoftFloat}.}

\subsubsection{SoftBLAS}
SoftBLAS is built on Berkeley SoftFloat~3e \citep{hauser_softfloat}.  Following SoftFloat, it requires a C99-compliant host and passes all floating-point quantities by value.  Every arithmetic primitive used by every routine is a SoftFloat call: there is no path to a hardware floating-point instruction, and therefore no path to hardware-dependent rounding, contraction, or intermediate precision.  For reasons enumerated by \citet{Davis2024}, this is the only current method which guarantees that the output of a floating-point computation is a pure function of its inputs, and therefore deterministic across hosts.

SoftBLAS covers the four \textsc{ieee} 754 binary precisions used across the stack:  half (\texttt{h}), single (\texttt{s}), double (\texttt{d}), and quadruple (\texttt{q}).\footnote{Complex routines are supplied at half (\texttt{i}), single (\texttt{c}), double (\texttt{z}), and quadruple (\texttt{v}) precisions.}  Routines follow the \textsc{blas} prefixing convention extended to the half and quad types.  Real-valued routines across all three levels are complete as of the cited release; complex-valued routines are in progress but sufficient to support Lagoon's requirements.

\begin{table}[h]
\centering
\caption{SoftBLAS coverage by level (real-valued).}
\label{tab:coverage}
\begin{tabular}{@{}lll@{}}
\toprule
Level & Cost & Representative routines \\
\midrule
1 & $O(n)$   & \texttt{?dot}, \texttt{?axpy}, \texttt{?scal}, \texttt{?nrm2}, \texttt{?asum}, etc. \\
2 & $O(n^2)$ & \texttt{?gemv} \\
3 & $O(n^3)$ & \texttt{?gemm} \\
\bottomrule
\end{tabular}
\end{table}

The rounding mode is a global, \texttt{softblas\_roundingMode} (a \texttt{char} typedef declared in \texttt{softblas.h}) interpreted by the underlying SoftFloat layer.\footnote{The transcendental function implementations ignore the rounding mode, which thus applies only to the arithmetic operations.}  The admissible values are the SoftFloat~3e rounding modes (round to nearest even, round toward zero, round toward $+\infty$, round toward $-\infty$, and round to nearest-away).\footnote{\textsc{ieee} 754-2008 introduced another mode, round to nearest ties to max magnitude, which is not included in Urbit's toolset.}  The default is round to nearest-even, \texttt{\%n}.  The Hoon specification names the rounding mode in the \texttt{\%base} distribution, and the jet is required to honor it.

SoftBLAS satisfies three conditions that conventional \textsc{blas} libraries do not:
\begin{enumerate}
  \item \emph{Fixed reduction order}.  Every reduction (dot products,
    \texttt{gemm} inner loops, norms) sums in one canonical order, identical to the order the Hoon specification uses.  No tiling or threading is permitted to reassociate it.  (Urbit runtimes are, in any case, single-threaded at the time of writing.)
  \item \emph{No contraction.} Multiply-add is two roundings, never a fused one, because the specification rounds twice.  There is no \textsc{fma} path, which as \citet{Davis2024} exhibited is prone to nondeterminism.
  \item \emph{Software rounding.} All rounding is performed by SoftFloat in software, so it is identical on every host regardless of native floating-point behavior.
\end{enumerate}

\subsubsection{SoftUnum}

SoftUnum is a software-defined universal number (unum) library that implements the 2022 posit standard~\citep{positstd2022}.  It supports five widths (8, 16, 32, 64, and 128 bits) with $es=2$ throughout.  The quire accumulator is implemented in software and is bit-identical across hosts.  The library is jetted into Lagoon's \texttt{\%unum} kind.

\section{Types}

The mathematical base types are enumerated in Listing~\ref{lst:types}.  These have occasioned the introduction of several new atom auras, given in Table~\ref{tab:auras}.  We have not introduced an aura for two's-complement integers at the current time, but these will likely nest under \texttt{@s} (signed integer) in the future.  The \texttt{@rqb} through \texttt{@rqq} auras are quire accumulators for the corresponding posit widths, and are intended for internal algorithms not general use.\footnote{A quire is a $16n$-bit two's-complement fixed-point accumulator with scale $2^{-qscale}$ ($qscale = 8n-16$): a posit value v contributes the exact integer $v * 2^qscale$.}

\begin{lstlisting}[style=listingblock,
                   label=lst:types,
                   caption={Mathematical base types.  From \texttt{/sur/lagoon}.}]
+$  kind              ::  $kind:  type of array scalars
  $?  %i754           ::  IEEE 754 float
      %uint           ::  unsigned integer
      %int2           ::  2s-complement integer (/lib/twoc)
      %unum           ::  unum/posit (/lib/unum)
      %cplx           ::  complex (/lib/complex)
      %fixp           ::  fixed-point Q a.b (/lib/fixed)
  ==
\end{lstlisting}

\begin{table}[h]
\centering
\caption{Existing and new atom auras.  From \texttt{/sur/lagoon}.}
\label{tab:auras}
\begin{tabular}{@{}rl@{}}
\toprule
Atom aura & Description \\
\midrule
\texttt{@rh  } & 16-bit \textsc{ieee} 754 float \\
\texttt{@rs  } & 32-bit \textsc{ieee} 754 float \\
\texttt{@rd  } & 64-bit \textsc{ieee} 754 float \\
\texttt{@rq  } & 128-bit \textsc{ieee} 754 float \\
\midrule
\texttt{@rpb } & 8-bit posit \\
\texttt{@rph } & 16-bit posit \\
\texttt{@rps } & 32-bit posit \\
\texttt{@rpd } & 64-bit posit \\
\texttt{@rpq } & 128-bit posit \\
\texttt{@rqbH} & 8-bit quire (128-bit accumulator for \texttt{@rpb}) \\
\texttt{@rqhI} & 16-bit quire (256-bit accumulator for \texttt{@rph}) \\
\texttt{@rqsJ} & 32-bit quire (512-bit accumulator for \texttt{@rps}) \\
\texttt{@rqdK} & 64-bit quire (1024-bit accumulator for \texttt{@rpd}) \\
\texttt{@rqqL} & 128-bit quire (2048-bit accumulator for \texttt{@rpq}) \\
\midrule
\texttt{@ch  } & 32-bit complex (two 16-bit floats) \\
\texttt{@cs  } & 64-bit complex (two 32-bit floats) \\
\texttt{@cd  } & 128-bit complex (two 64-bit floats) \\
\texttt{@cq  } & 256-bit complex (two 128-bit floats) \\
\bottomrule
\end{tabular}
\end{table}

Aura types have been named after the target width in the cases of complex numbers and quires.

\subsection{Verification}

The determinism architecture reduces correctness to one question per routine:  does the jet equal the specification on every input? We answer it insofar as possible with oracles external to both the Hoon and C implementations.  SoftFloat by and large served as the oracle for the \textsc{ieee} 754 floating-point operations (real and complex), while the posit mathematics were verified against two independent references:  SoftPosit~\citep{leong_softposit} (\texttt{posit8}, \texttt{posit16}, \texttt{posit32}) and the Universal library~\citep{omtzigt2020universal} (\texttt{posit64}, \texttt{posit128}).

The quire's defining property---exact accumulation until a single final rounding---is checked directly against an exact rational sum.

Elementary functions are checked against \textsc{mpfr}~\citep{fousse2007mpfr} at high precision.  At half precision the input space is $2^{16}$ values and the test is exhaustive; at single precision the $2^{32}$ space is likewise tested in full.  At double and quadruple precision, where exhaustive testing is infeasible, we report maximum observed unit in the last place (\textsc{ulp}) error over a sampled and edge-weighted input set rather than a guarantee.  We claim a faithful-rounding ($\leq 1$~\textsc{ulp}) result where the testing supports it and a measured bound where it does not.

\section{Contributions}

This deterministic numerical stack contributes several properties to the broader discussion of numerical determinism and reproducibility.  Determinism is a structural property of Urbit:  while reproducible floating point is a standing goal of several parties (\citet{demmel2013reproducible}; \citet{intel_cnr}), Urbit contributes a philosophical first-class enforcement mechanism.\footnote{Jet co-design methodology is treated separately; see \patp{lagrev-nocfep}, pp.~TODO–TODO in this volume, for a companion article.}

Although not novel, the library covers a broad range of bit widths and precisions, including half-precision and quadruple-precision, which are not widely supported in other numerical computing libraries.  Likewise we provided faithfully-rounded elementary functions at all precisions.

Lagoon's heterogeneous-kind array design with a single \texttt{\$ray} type is a novel approach to array libraries, allowing for in-band kind dispatch and fall-through to specification.  As mentioned previously, this permits more flexibility and extensibility in the base representation than NumPy, Julia, or the \textsc{apl}/J/K tradition.  The \texttt{kind} field allows for easy addition of new numeric types without requiring changes to the underlying array structure, and the fall-through mechanism ensures that un-jetted kinds can still be computed correctly using the Hoon specification.

\section{Benchmarks}

While a software-defined floating-point library will never match the performance of a hardware-accelerated one, we have nevertheless seen impressive speedups by refining the algorithms and the runtime jets.  The benchmarks reported here are for a single-threaded, single-core execution on a modern 64-bit compiler \textsc{cpu} (AArch64, Apple M4 Max).  They should serve as order-of-magnitude indications on contemporary hardware.

The first benchmark results tabulate what the jetted system buys over interpreting the Hoon specification directly.  Table~\ref{tab:jet} reports jetted versus interpreted time for \textsc{ieee} 754 \texttt{/lib/math}.  For most trials, 100,000 runs were evaluated and averaged.  For the very slowest functions (e.g. half-precision Taylor series inverse sine), this number was modulated downwards substantially.\footnote{Exhaustive results for all precisions are available at \href{https://github.com/urbit/numerics/tree/main/benchmark}{\texttt{urbit/numerics@main:benchmark}}.}  The original \texttt{/lib/math} and Saloon implementations of transcendentals were na\"{i}ve Taylor series expansions and serve as a worst-case demonstration.  Accuracies are comparable for the number of Taylor series terms utilized to achieve an appropriate relative tolerance, but the Chebyshev basis functions are uniformly faster and more correct.  All values reported are measured within a running Urbit ship (rather than from an independent test harness).

\begin{landscape}
\begin{table}[h]
\centering
\caption{Elapsed times and relative slowdowns of \textsc{ieee} 754 jetted Chebyshev functions for double-precision \texttt{@rd} atom operations with \texttt{/lib/math}, interpreted Chebyshev functions, and interpreted Taylor series functions.  Slowdown is reported relative to the fastest case, the jetted Chebyshev functions.}
\label{tab:jet}
\begin{tabular}{lrrrrr}
\toprule
Arm & Chebyshev (jet) & Chebyshev (int.) & Taylor (int.) & C.\ slowdown & T.\ slowdown \\
\midrule
\texttt{+exp} & 1.67 µs & 296 µs & 11.0 ms & 177× & 6,565× \\
\texttt{+log} & 1.76 µs & 164 µs & 13.3 ms & 94× & 7,600× \\
\texttt{+sin} & 1.95 µs & 278 µs & 0.94 ms & 143× & 480× \\
\texttt{+cos} & 1.97 µs & 275 µs & 0.94 ms & 140× & 476× \\
\texttt{+tan} & 4.53 µs & 241 µs & 1.30 ms & 53× & 288× \\
\texttt{+atan} & 2.23 µs & 47.6 µs & 109 ms & 21× & 48,784× \\
\texttt{+atan2} & 2.72 µs & 56.2 µs & 11.5 ms & 21× & 4,237× \\
\texttt{+asin} & 2.07 µs & 54.8 µs & 205 ms & 27× & 99,130× \\
\texttt{+acos} & 2.10 µs & 51.2 µs & 5.17 ms & 24× & 2,459× \\
\texttt{+sqt} & 1.92 µs & 10.3 µs & 73.5 µs & 5× & 38× \\
\texttt{+cbt} & 2.44 µs & 467 µs & 111 ms & 192× & 45,723× \\
\texttt{+pow} & 2.87 µs & 656 µs & 1.68 ms & 229× & 585× \\
\texttt{+pow-n} & 2.62 µs & 217 µs & 219 µs & 83× & 83× \\
\texttt{+log-2} & 2.34 µs & 164 µs & 13.3 ms & 70× & 5,668× \\
\texttt{+log-10} & 2.38 µs & 165 µs & 13.4 ms & 69× & 5,638× \\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

The \texttt{/lib/unum} transcendentals are currently implemented as Chebyshev basis expansions like the \texttt{/lib/math} operations.  Results are included in Table~\ref{tab:unum}.

\begin{landscape}
\begin{table}[h]
\centering
\caption{Elapsed times and relative slowdowns of unum jetted Chebyshev functions for 32-bit \texttt{@rps} \texttt{posit32} operations with \texttt{/lib/unum}, interpreted Chebyshev functions, jetted Taylor series functions, and interpreted Taylor series functions.  Slowdown is reported relative to the jetted Chebyshev functions.}
\label{tab:unum}
\begin{tabular}{lrrrrrr}
\toprule
Arm & Chebyshev (jet) & Chebyshev (int.) & Taylor (jet) & Taylor (int.) & C.\ slowdown & T.\ slowdown \\
\midrule
% \texttt{+add} & 1.75 µs & 232 µs & 1.69 µs & 374 µs & 132× & 213× \\
% \texttt{+sub} & 1.73 µs & 245 µs & 1.76 µs & 340 µs & 142× & 197× \\
% \texttt{+mul} & 1.64 µs & 217 µs & 1.62 µs & 277 µs & 132× & 169× \\
% \texttt{+div} & 1.43 µs & 242 µs & 1.44 µs & 344 µs & 169× & 241× \\
% \texttt{+fma} & 1.76 µs & 308 µs & 1.90 µs & 493 µs & 175× & 99× \\
% \texttt{+neg} & 732 ns & 4.72 µs & 735 ns & 19.7 µs & 6× & 27× \\
% \texttt{+lth} & 1.30 µs & 8.98 µs & 1.28 µs & 20.8 µs & 7× & 16× \\
\texttt{+exp} & 16.5 µs & 550 µs & 25.3 µs & 19.4 ms & 33× & 1,176× \\
\texttt{+log} & 17.9 µs & 739 µs & 39.9 µs & 39.1 ms & 41× & 2,184× \\
\texttt{+sin} & 18.3 µs & 813 µs & 34.8 µs & 32.3 ms & 44× & 1,765× \\
\texttt{+cos} & 18.3 µs & 817 µs & 35.7 µs & 32.5 ms & 45× & 1,775× \\
\texttt{+atan} & 18.3 µs & 832 µs & 850.3 µs & 63.7 ms & 45× & 3,480× \\
% \texttt{+atan2} & 2.72 µs & 56.2 µs & 11.5 ms & 21× & 4,237× \\
\texttt{+asin} & 39.6 µs & 2.01 ms & --- & --- & 50× & --- \\
\texttt{+acos} & 40.9 µs & 2.15 ms  & --- & --- & 52× & --- \\
\texttt{+sqt} & 17.1 µs & 209 µs & 17.3 µs & 341 µs & 12× & 20× \\
\texttt{+cbt} & 32.0 µs & 1.76 ms  & --- & --- & 55× & --- \\
\texttt{+pow} & 31.2 µs & 1.40 ms & 60.7 µs & 56.9 ms & 45× & 1,824× \\
\texttt{+pow-n} & 2.81 µs & 698 µs  & --- & --- & 248× & --- \\
\texttt{+log-2} & 18.1 µs & 743 µs & --- & --- & 41× & ---\\
\texttt{+log-10} & 18.1 µs & 756 µs & --- & --- & 42× & --- \\
\\
\texttt{+fdp} & 259 ns & 157 µs & 270 ns & 205 µs & 606× & --- \\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

\subsection{Transcendental accuracy}

TODO
% This subsection must carry the measured max-ULP table: exhaustive at f16/f32 (you already say you did this), sampled at f64/f128.

\sloppy
The libraries instrumented with the Chebyshev basis produce ``faithfully rounded functions'' rather than ``correctly rounded functions''.  The distinction is that a faithfully rounded result is within 1~\textsc{ulp} of the exact result \citep{Rump2008}, while a correctly rounded result is the exact representable value closest to the exact result.  Faithful rounding is often sufficient for numerical applications and is generally faster to compute than correct rounding.  Truly correct rounding is often prohibitively expensive, especially for transcendental functions, and can require much higher intermediate precision to attain.\footnote{See \textsc{cr-libm} \citep{daramy2003crlibm}, \textsc{rlibm}~\citep{lim2021rlibm}, and the \textsc{core-math} project~\citep{coremath} for further details.}

\subsection{The price of determinism}

The third field of benchmarking measures the overhead of the determinism guarantee relative to a conventional, nondeterministic, hardware \textsc{blas}. We report SoftBLAS against a stock optimized \textsc{blas} on the same host. We expect SoftBLAS to be slower; here we quantify \emph{how much} (our implementation of) determinism costs.

TODO

\section{Future Work}

We have focused more than our taste on \textsc{ieee} 754 given its ongoing dominance in numerical computation, but we anticipate the libraries beyond \texttt{/lib/math} to grow in relative importance in the future.

\sloppy
The complex-valued SoftBLAS routines are in progress and will complete the Level 1--3 coverage.  Quire operations in \texttt{/lib/unum} are implemented within the generic posit core but are not yet surfaced as callable top-level arms; exposing them is a prerequisite both for user access and for routing them through the verification harness.  The unum specification also defines valids, which are arithmetic intervals which have not yet seen widespread adoption.

Saloon will grow in methods adjacent to scientific computing:  conjugate gradient minimization and other optimization algorithms; random number generators; and the like.  The fixed-point library is relatively under\-developed compared to the other arithmetics, and will be expanded to support more general Q-format widths and operations.  More components of the various libraries need to be upgraded to a state-of-the-art Chebyshev basis and benchmarked for performance.

Finally, Maroon, the machine learning library, is in its infancy and will be developed to support a range of machine learning algorithms and techniques.  Part of this field of consideration encompasses the utilization of \textsc{gpu} programming for \textsc{llm}-based computing within or adjacent to Urbit.  \textsc{gpu} numerical nondeterminism and host portability presents a challenge of the same order of magnitude as the baseline reproducibility we have tackled over the past several years in this project.  One can envision a ReproCL as a deterministic counterpart to OpenCL, but it is unclear whether sufficient warrant exists to justify the effort as of the mid-2020s.  Heterogeneous compute would be benefitted by the adoption of alternatives to \textsc{ieee} 754 with better guarantees of portability:  not just unums but \texttt{bfloat16} and other newer types may answer the need.

\section{Conclusion}

Urbit as a deterministic state machine cannot treat numerical determinism or bitwise reproducibility as merely a tolerance.  The architectural stack we have described here constitutes a partial answer to the challenge underlying \citet{Davis2024}:  how can a numerical computation built on a particular stack be reliably reproducible?  We maintain this determinism by painstaking construction.  The computational cost of determinism is currently (and regrettably) non-negligible, while one can hope for future developments, protocols, and architectures that make it more natural and tenable.\tombstone

\selectlanguage{USenglish}
\printbibliography
% TODO check all references for reality
\end{document}
